---
title: "Foreign Trade and Balance of Payments"
subtitle: "Inter-Table Relations — Based on Choudhari Chapter"
author: "Economics Analysis"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    toc-location: left
    number-sections: true
    code-fold: true
    code-summary: "Show Code"
    smooth-scroll: true
    page-layout: full
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
editor:
  markdown:
    wrap: sentence
---

```{r}
#| label: setup
#| include: false

library(dplyr)
library(kableExtra)
library(knitr)
```

# Overview

This document presents four inter-related tables drawn from the **Foreign Trade and Balance of Payments** chapter (Choudhari).
Together they trace how India's **Balance of Trade** feeds into the **Current Account**, how the **Capital Account** is built up, and how both accounts combine into the **Overall Balance of Payments (BOP)**.

The data below are illustrative figures (in ₹ Crore) consistent with the textbook treatment of the topic.

---

# Table 1 — Balance of Trade (BoT)

The Balance of Trade records the difference between the value of **merchandise exports** and **merchandise imports** over a given period.

$$\text{Trade Balance} = \text{Exports (X)} - \text{Imports (M)}$$

A *negative* value indicates a **trade deficit** (imports exceed exports).

```{r}
#| label: table1-bot
#| echo: true

bot <- data.frame(
  Year    = c("2017–18", "2018–19", "2019–20", "2020–21", "2021–22"),
  Exports = c(1852600, 2310600, 2113200, 1756400, 3218700),
  Imports = c(2887800, 3459400, 3001300, 2398900, 4899100),
  Trade_Balance = c(-1035200, -1148800, -888100, -642500, -1680400)
)

bot |>
  kbl(
    caption   = "Table 1: India's Balance of Trade (₹ Crore)",
    booktabs  = TRUE,
    col.names = c("Year", "Exports (X)", "Imports (M)", "Trade Balance (X − M)"),
    format.args = list(big.mark = ",")
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = TRUE
  ) |>
  column_spec(4, bold = TRUE,
              color = ifelse(bot$Trade_Balance < 0, "red", "darkgreen"))
```

> **Key relation:** The *Trade Balance* column from Table 1 is carried forward directly into Table 2 as the starting component of the Current Account.

---

# Table 2 — Current Account Balance

The Current Account has three sub-components:

| Sub-component | Description |
|---|---|
| **Trade Balance** | Goods exports minus goods imports (= Table 1, col 4) |
| **Net Invisibles** | Net receipts from services, remittances & transfers |
| **Current Account Balance** | Trade Balance + Net Invisibles |

$$\text{Current Account Balance (CAB)} = \text{Trade Balance} + \text{Net Invisibles}$$

```{r}
#| label: table2-ca
#| echo: true

current_account <- data.frame(
  Year            = bot$Year,
  Trade_Balance   = bot$Trade_Balance,           # ← linked from Table 1
  Net_Invisibles  = c(518600, 601400, 647300, 712800, 789200),
  CA_Balance      = NA_real_
)

current_account <- current_account |>
  mutate(CA_Balance = Trade_Balance + Net_Invisibles)

current_account |>
  kbl(
    caption   = "Table 2: India's Current Account Balance (₹ Crore)",
    booktabs  = TRUE,
    col.names = c("Year", "Trade Balance (from T1)", "Net Invisibles",
                  "Current Account Balance"),
    format.args = list(big.mark = ",")
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = TRUE
  ) |>
  column_spec(2, italic = TRUE, color = "grey40") |>
  column_spec(4, bold = TRUE,
              color = ifelse(current_account$CA_Balance < 0, "red", "darkgreen"))
```

> **Key relation:** The *Current Account Balance* from Table 2 is one of the two main inputs into the Overall BOP (Table 4). The *Trade Balance* column is identical to Table 1 column 4 — demonstrating a direct link between the two tables.

---

# Table 3 — Capital Account Balance

The Capital Account records financial flows between India and the rest of the world:

| Component | Description |
|---|---|
| **FDI (net)** | Foreign Direct Investment inflows minus outflows |
| **FPI / FII (net)** | Foreign Portfolio Investment (equity + debt) |
| **External Borrowings** | ECB, NRI deposits, and other loans |
| **Capital Account Balance** | Sum of all capital flow items |

$$\text{Capital Account Balance (KAB)} = \text{FDI} + \text{FPI} + \text{External Borrowings}$$

```{r}
#| label: table3-ka
#| echo: true

capital_account <- data.frame(
  Year                = bot$Year,
  FDI_Net             = c(226700, 305400, 432100, 486300, 584800),
  FPI_Net             = c( 25400,  -6800, 115200, 272100, -128600),
  External_Borrowings = c(180500, 195300,  88400, 115600,  237800),
  KA_Balance          = NA_real_
)

capital_account <- capital_account |>
  mutate(KA_Balance = FDI_Net + FPI_Net + External_Borrowings)

capital_account |>
  kbl(
    caption   = "Table 3: India's Capital Account Balance (₹ Crore)",
    booktabs  = TRUE,
    col.names = c("Year", "FDI (net)", "FPI/FII (net)",
                  "External Borrowings", "Capital Account Balance"),
    format.args = list(big.mark = ",")
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = TRUE
  ) |>
  column_spec(5, bold = TRUE,
              color = ifelse(capital_account$KA_Balance >= 0, "darkgreen", "red"))
```

> **Key relation:** The *Capital Account Balance* from Table 3 is the second main input into the Overall BOP (Table 4).

---

# Table 4 — Overall Balance of Payments (BOP)

The **Balance of Payments** consolidates both the Current Account and the Capital Account.
Any surplus or deficit is settled through a change in **Foreign Exchange Reserves**.

$$\text{Overall BOP} = \text{CAB (Table 2)} + \text{KAB (Table 3)}$$

$$\text{Change in Forex Reserves} = -\,\text{Overall BOP}$$

*(A positive Overall BOP means reserves *increase*; a negative BOP means reserves *decrease*.)*

```{r}
#| label: table4-bop
#| echo: true

bop <- data.frame(
  Year       = bot$Year,
  CA_Balance = current_account$CA_Balance,   # ← linked from Table 2
  KA_Balance = capital_account$KA_Balance,   # ← linked from Table 3
  Overall_BOP = NA_real_,
  Change_Reserves = NA_real_
)

bop <- bop |>
  mutate(
    Overall_BOP     = CA_Balance + KA_Balance,
    Change_Reserves = -Overall_BOP
  )

bop |>
  kbl(
    caption   = "Table 4: India's Overall Balance of Payments (₹ Crore)",
    booktabs  = TRUE,
    col.names = c("Year", "CAB (from T2)", "KAB (from T3)",
                  "Overall BOP", "Change in Forex Reserves"),
    format.args = list(big.mark = ",")
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = TRUE
  ) |>
  column_spec(2, italic = TRUE, color = "grey40") |>
  column_spec(3, italic = TRUE, color = "grey40") |>
  column_spec(4, bold = TRUE,
              color = ifelse(bop$Overall_BOP >= 0, "darkgreen", "red")) |>
  column_spec(5, bold = TRUE,
              color = ifelse(bop$Change_Reserves >= 0, "darkgreen", "red"))
```

---

# Relations Between the Four Tables

The diagram below summarises how each table feeds into the next.

```{r}
#| label: relation-summary
#| echo: false

relation <- data.frame(
  Step = c(
    "T1 → T2",
    "T2 → T4",
    "T3 → T4",
    "T4 (derived)"
  ),
  From_Table = c(
    "Table 1 — Balance of Trade",
    "Table 2 — Current Account",
    "Table 3 — Capital Account",
    "Table 4 — Overall BOP"
  ),
  Column_Transferred = c(
    "Trade Balance (X − M)",
    "Current Account Balance",
    "Capital Account Balance",
    "Change in Forex Reserves"
  ),
  Used_As = c(
    "First component of Current Account (Table 2)",
    "First input to Overall BOP (Table 4)",
    "Second input to Overall BOP (Table 4)",
    "Derived as −(CAB + KAB)"
  )
)

relation |>
  kbl(
    caption   = "Summary of Inter-Table Relations",
    booktabs  = TRUE,
    col.names = c("Link", "Source Table", "Column / Value Transferred", "How It Is Used")
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = TRUE
  ) |>
  column_spec(1, bold = TRUE, width = "6em") |>
  column_spec(2, width = "18em") |>
  column_spec(3, width = "20em") |>
  column_spec(4, width = "24em")
```

## Verification: Consistency Check Across Tables

```{r}
#| label: verification
#| echo: true

# Confirm that CA_Balance in Table 4 equals the one computed in Table 2
stopifnot(all(bop$CA_Balance == current_account$CA_Balance))

# Confirm that KA_Balance in Table 4 equals the one computed in Table 3
stopifnot(all(bop$KA_Balance == capital_account$KA_Balance))

# Confirm Trade Balance in Table 2 equals the one computed in Table 1
stopifnot(all(current_account$Trade_Balance == bot$Trade_Balance))

message("All inter-table relations verified — values are consistent across Table 1, 2, 3, and 4.")
```

---

# Conceptual Summary

| Account | What it measures | Linked tables |
|---|---|---|
| **Balance of Trade** | Goods exports vs imports | Table 1 → Table 2 |
| **Current Account** | Goods + Services + Transfers | Table 2 → Table 4 |
| **Capital Account** | Financial flows (FDI, FPI, Loans) | Table 3 → Table 4 |
| **Overall BOP** | Current + Capital accounts | Table 4 (final) |

> *"The Balance of Payments always balances — any Current Account deficit must be financed by a surplus on the Capital Account or by drawing down Foreign Exchange Reserves."*  
> — Choudhari, *Foreign Trade and Balance of Payments*
