---
title: "Financial Dependence and Healthcare Utilization among Older Adults"
subtitle: "Evidence from the Longitudinal Ageing Study in India (LASI) Wave 1, 2017–18"
author: "Research Analysis"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    toc-location: left
    number-sections: true
    code-fold: true
    code-summary: "Show Code"
    code-tools: true
    smooth-scroll: true
    page-layout: full
    fig-width: 10
    fig-height: 7
    fig-dpi: 150
    lightbox: true
    embed-resources: true
  pdf:
    documentclass: article
    toc: true
    toc-depth: 3
    number-sections: true
    fig-width: 10
    fig-height: 7
    fig-pos: "H"
    geometry:
      - top=25mm
      - bottom=25mm
      - left=20mm
      - right=20mm
    fontsize: 11pt
    colorlinks: true
    keep-tex: true
    pdf-engine: xelatex
    include-in-header:
      text: |
        \usepackage{longtable}
        \usepackage{booktabs}
        \usepackage{array}
        \usepackage{multirow}
        \usepackage{colortbl}
        \usepackage{xcolor}
        \usepackage{float}
        \usepackage{graphicx}
        \usepackage{fancyhdr}
        \usepackage{enumitem}
        \usepackage{amsmath}
        \usepackage{amssymb}
        \usepackage{geometry}
        \usepackage{setspace}
        \usepackage{parskip}
        \usepackage{hyperref}
        \usepackage{lscape}
        \usepackage{rotating}
        \usepackage{siunitx}
        \usepackage{pdflscape}
execute:
  warning: false
  message: false
  echo: false
editor: 
  markdown: 
    wrap: sentence
---

```{r}
#| label: latex-setup
#| include: false
```

```{r}
#| label: css-fix
#| results: asis
#| echo: false

if (!knitr::is_latex_output()) {
  cat('
<style>
table {
  width: 100% !important;
}
</style>
')
}
```

```{r}
#| label: setup
#| include: false

library(haven)
library(dplyr)
library(stargazer)
library(kableExtra)
library(knitr)

options(knitr.table.format = "latex")
options(kableExtra.latex.load_packages = FALSE)

HH_FI <- read_dta("HH_FI.dta")

HH_FI <- HH_FI |>
  mutate(across(where(haven::is.labelled), as.numeric))

HH_FI <- HH_FI |>
  mutate(
    gender_label    = factor(sex, levels = c(0, 1), labels = c("Female", "Male")),
    zone_label      = factor(zone),
    residence_label = factor(residence_type, levels = c(0, 1), labels = c("Urban", "Rural"))
  )

stargazer_output_type <- if (knitr::is_latex_output()) "latex" else "html"
```

```{r}
#| echo: true
#| code-fold: true
#| label: variable-groups
#| include: false

dependent_variables <- c(
  "healthcare",
  "inpatient_hospitalization",
  "total_facilities_visited",
  "hospitalisation_count"
)

independent_variables <- c(
  "health_insurance_covered",
  "pension_status",
  "financial_support"
)

continuous_control_variables <- c(
  "age",
  "sex",
  "residence_type"
)

religion_labels <- c(
  "1" = "Hindu", "2" = "Muslim", "3" = "Christian",
  "4" = "Sikh", "5" = "Buddhist", "6" = "Jain",
  "9" = "Other", "10" = "None"
)

caste_labels <- c(
  "1" = "SC", "2" = "ST", "3" = "OBC", "4" = "General"
)

marital_labels <- c(
  "1" = "Married", "2" = "Widowed", "3" = "Divorced",
  "4" = "Separated", "5" = "Never married",
  "6" = "Deserted", "7" = "Other"
)

category_label_map <- list(
  religion           = religion_labels,
  caste_social_group = caste_labels,
  marital_status     = marital_labels
)

categorical_control_variables <- names(category_label_map)
```

```{r}
#| label: table-functions
#| include: false

expand_categorical_rows <- function(variable_name, label_map, data) {
  values      <- as.numeric(data[[variable_name]])
  non_missing <- values[!is.na(values)]
  all_vals    <- sort(unique(non_missing))

  header_row <- data.frame(
    Variable = variable_name,
    N        = "",
    Mean     = "",
    SD       = "",
    Min      = "",
    Max      = "",
    is_header = TRUE,
    stringsAsFactors = FALSE
  )

  sub_rows <- lapply(seq_along(all_vals), function(i) {
    val     <- all_vals[i]
    label   <- ifelse(as.character(val) %in% names(label_map),
                      label_map[as.character(val)], as.character(val))
    dummy   <- as.numeric(non_missing == val)
    n_obs   <- sum(non_missing == val, na.rm = TRUE)
    data.frame(
      Variable  = paste0("  ", i, ". ", label),
      N         = as.character(n_obs),
      Mean      = as.character(round(mean(dummy, na.rm = TRUE), 3)),
      SD        = as.character(round(sd(dummy,   na.rm = TRUE), 3)),
      Min       = as.character(round(min(dummy,  na.rm = TRUE), 3)),
      Max       = as.character(round(max(dummy,  na.rm = TRUE), 3)),
      is_header = FALSE,
      stringsAsFactors = FALSE
    )
  })

  do.call(rbind, c(list(header_row), sub_rows))
}

make_overall_control_table <- function(cont_vars, cat_vars,
                                       label_map_list, table_title) {
  cont_rows <- lapply(cont_vars, function(one_var) {
    values      <- as.numeric(HH_FI[[one_var]])
    non_missing <- values[!is.na(values)]
    data.frame(
      Variable  = one_var,
      N         = as.character(length(non_missing)),
      Mean      = as.character(round(mean(non_missing, na.rm = TRUE), 3)),
      SD        = as.character(round(sd(non_missing,   na.rm = TRUE), 3)),
      Min       = as.character(round(min(non_missing,  na.rm = TRUE), 3)),
      Max       = as.character(round(max(non_missing,  na.rm = TRUE), 3)),
      is_header = FALSE,
      stringsAsFactors = FALSE
    )
  })

  cat_rows <- lapply(cat_vars, function(one_var) {
    expand_categorical_rows(one_var, label_map_list[[one_var]], HH_FI)
  })

  all_rows     <- do.call(rbind, c(cont_rows, cat_rows))
  header_idx   <- which(all_rows$is_header)
  display_table <- all_rows[, c("Variable","N","Mean","SD","Min","Max")]

  tbl <- display_table |>
    kableExtra::kbl(
      caption   = table_title,
      booktabs  = TRUE,
      longtable = TRUE,
      linesep   = "",
      col.names = c("Variable", "N", "Mean", "SD", "Min", "Max")
    ) |>
    kableExtra::kable_styling(
      latex_options = c("hold_position", "repeat_header", "striped", "scale_down"),
      full_width    = TRUE,
      font_size     = 9
    ) |>
    kableExtra::column_spec(1, bold = FALSE, width = "5cm")

  for (i in header_idx) {
    tbl <- tbl |>
      kableExtra::row_spec(i, bold = TRUE, background = "#f0f0f0")
  }

  tbl
}

make_wide_subgroup_table <- function(variable_list, grouping_column, table_title) {
  all_groups <- sort(unique(na.omit(HH_FI[[grouping_column]])))

  all_rows <- lapply(variable_list, function(one_var) {
    stats_across_groups <- lapply(all_groups, function(one_group) {
      in_group    <- HH_FI[[grouping_column]] == one_group &
                     !is.na(HH_FI[[grouping_column]])
      values      <- as.numeric(HH_FI[[one_var]])[in_group]
      non_missing <- values[!is.na(values)]
      data.frame(
        N    = length(non_missing),
        Mean = as.character(round(mean(non_missing, na.rm = TRUE), 3)),
        SD   = as.character(round(sd(non_missing,   na.rm = TRUE), 3)),
        Min  = as.character(round(min(non_missing,  na.rm = TRUE), 3)),
        Max  = as.character(round(max(non_missing,  na.rm = TRUE), 3)),
        stringsAsFactors = FALSE
      )
    })
    wide_row          <- do.call(cbind, stats_across_groups)
    wide_row$Variable <- one_var
    wide_row
  })

  result_table    <- do.call(rbind, all_rows)
  group_col_names <- unlist(lapply(all_groups, function(g) {
    c(paste0(g,"_N"), paste0(g,"_Mean"), paste0(g,"_SD"),
      paste0(g,"_Min"), paste0(g,"_Max"))
  }))
  colnames(result_table) <- c(group_col_names, "Variable")
  result_table           <- result_table[, c("Variable", group_col_names)]

  header_above <- c(" " = 1)
  for (g in all_groups) { header_above[as.character(g)] <- 5 }

  result_table |>
    kableExtra::kbl(
      caption   = table_title,
      booktabs  = TRUE,
      longtable = TRUE,
      linesep   = "",
      col.names = c("Variable",
                    rep(c("N", "Mean", "SD", "Min", "Max"), length(all_groups)))
    ) |>
    kableExtra::add_header_above(header_above) |>
    kableExtra::kable_styling(
      latex_options = c("hold_position", "repeat_header", "striped", "scale_down"),
      full_width    = TRUE,
      font_size     = 9
    ) |>
    kableExtra::column_spec(1, bold = TRUE, width = "4cm")
}

make_wide_categorical_subgroup_table <- function(cat_vars, label_map_list,
                                                  grouping_column, table_title) {
  all_groups <- sort(unique(na.omit(HH_FI[[grouping_column]])))

  all_rows <- lapply(cat_vars, function(one_var) {
    values   <- as.numeric(HH_FI[[one_var]])
    all_vals <- sort(unique(values[!is.na(values)]))
    label_map <- label_map_list[[one_var]]

    # Parent header row — blank stats across all groups
    n_stat_cols <- length(all_groups) * 5
    header_vals <- as.list(rep("", n_stat_cols))
    header_row  <- data.frame(
      Variable  = one_var,
      is_header = TRUE,
      stringsAsFactors = FALSE
    )
    for (g in all_groups) {
      header_row[[paste0(g,"_N")]]    <- ""
      header_row[[paste0(g,"_Mean")]] <- ""
      header_row[[paste0(g,"_SD")]]   <- ""
      header_row[[paste0(g,"_Min")]]  <- ""
      header_row[[paste0(g,"_Max")]]  <- ""
    }

    # Sub-rows: one per category
    sub_rows <- lapply(seq_along(all_vals), function(i) {
      val   <- all_vals[i]
      label <- ifelse(as.character(val) %in% names(label_map),
                      label_map[as.character(val)], as.character(val))
      row <- data.frame(
        Variable  = paste0("  ", i, ". ", label),
        is_header = FALSE,
        stringsAsFactors = FALSE
      )
      for (one_group in all_groups) {
        in_group    <- HH_FI[[grouping_column]] == one_group &
                       !is.na(HH_FI[[grouping_column]])
        grp_values  <- as.numeric(HH_FI[[one_var]])[in_group]
        non_missing <- grp_values[!is.na(grp_values)]
        dummy       <- as.numeric(non_missing == val)
        row[[paste0(one_group,"_N")]]    <- as.character(sum(non_missing == val, na.rm = TRUE))
        row[[paste0(one_group,"_Mean")]] <- as.character(round(mean(dummy), 3))
        row[[paste0(one_group,"_SD")]]   <- as.character(round(sd(dummy),   3))
        row[[paste0(one_group,"_Min")]]  <- as.character(round(min(dummy),  3))
        row[[paste0(one_group,"_Max")]]  <- as.character(round(max(dummy),  3))
      }
      row
    })

    do.call(rbind, c(list(header_row), sub_rows))
  })

  result_table <- do.call(rbind, all_rows)
  header_idx   <- which(result_table$is_header)

  group_col_names <- unlist(lapply(all_groups, function(g) {
    c(paste0(g,"_N"), paste0(g,"_Mean"), paste0(g,"_SD"),
      paste0(g,"_Min"), paste0(g,"_Max"))
  }))

  display_table <- result_table[, c("Variable", group_col_names)]

  header_above <- c(" " = 1)
  for (g in all_groups) { header_above[as.character(g)] <- 5 }

  tbl <- display_table |>
    kableExtra::kbl(
      caption   = table_title,
      booktabs  = TRUE,
      longtable = TRUE,
      linesep   = "",
      col.names = c("Variable",
                    rep(c("N", "Mean", "SD", "Min", "Max"), length(all_groups)))
    ) |>
    kableExtra::add_header_above(header_above) |>
    kableExtra::kable_styling(
      latex_options = c("hold_position", "repeat_header", "striped", "scale_down"),
      full_width    = TRUE,
      font_size     = 9
    ) |>
    kableExtra::column_spec(1, bold = FALSE, width = "4cm")

  for (i in header_idx) {
    tbl <- tbl |>
      kableExtra::row_spec(i, bold = TRUE, background = "#f0f0f0")
  }

  tbl
}
```

# Data and Variables

## Data Source

This study uses Wave 1 of the Longitudinal Ageing Study in India (LASI), conducted in 2017--18.
LASI is a nationally representative longitudinal survey covering adults aged 45 and above across all Indian states and union territories.
The analysis uses the individual-level dataset, restricting the sample to respondents aged 45 years and above.

## Variable Description

### Dependent Variables

Four outcome variables are used.
**Healthcare** is a binary indicator equal to 1 if the individual visited any healthcare facility in the past 12 months.
**Inpatient hospitalization** is a binary indicator equal to 1 if the individual was admitted to a hospital.


### Independent Variables

Three key predictors capture financial protection.
**Health insurance coverage** is a binary indicator equal to 1 if the individual is covered by any health insurance scheme.
**Pension status** is a binary indicator equal to 1 if the individual receives any pension income.
**Financial support** is a binary indicator equal to 1 if the individual receives financial support from family or others.

### Control Variables

The models control for **age** in years, **sex** (1 = Male, 0 = Female), and **residence type** (1 = Rural, 0 = Urban).
**Religion**, **caste group**, and **marital status** are multi-category variables.
Each category is treated as a binary dummy variable and reported separately with its own N, Mean, SD, Min, and Max.

# Summary Statistics

## Overall Summary Statistics

### Dependent Variables

```{r}
#| echo: true
#| code-fold: true
#| label: overall-dependent
#| results: asis

stargazer(
  as.data.frame(HH_FI[, dependent_variables[1:2]]),
  type         = stargazer_output_type,
  title        = "Overall Summary Statistics — Dependent Variables",
  digits       = 3,
  summary.stat = c("n", "mean", "sd", "min", "max"),
  header       = FALSE,
  font.size    = "normalsize",
  table.placement = "H",
  label        = "tab:dep"
)
```

### Independent Variables

```{r}
#| echo: true
#| code-fold: true
#| label: overall-independent
#| results: asis

stargazer(
  as.data.frame(HH_FI[, independent_variables]),
  type         = stargazer_output_type,
  title        = "Overall Summary Statistics — Independent Variables",
  digits       = 3,
  summary.stat = c("n", "mean", "sd", "min", "max"),
  header       = FALSE,
  font.size    = "normalsize",
  table.placement = "H",
  label        = "tab:indep"
)
```

### Control Variables

```{r}
#| echo: true
#| code-fold: true
#| label: overall-control
make_overall_control_table(
  cont_vars      = continuous_control_variables,
  cat_vars       = categorical_control_variables,
  label_map_list = category_label_map,
  table_title    = "Overall Summary Statistics — Control Variables"
)
```

## Subgroup Comparisons

### By Gender

#### Dependent Variables

```{r}
#| echo: true
#| code-fold: true
#| label: gender-dependent
make_wide_subgroup_table(dependent_variables[1:2], "gender_label",
                         "Dependent Variables — By Gender")
```

#### Independent Variables

```{r}
#| echo: true
#| code-fold: true
#| label: gender-independent
make_wide_subgroup_table(independent_variables, "gender_label",
                         "Independent Variables — By Gender")
```

#### Control Variables

```{r}
#| echo: true
#| code-fold: true
#| label: gender-control-continuous
make_wide_subgroup_table(continuous_control_variables, "gender_label",
                         "Control Variables — By Gender")
```

```{r}
#| echo: true
#| code-fold: true
#| label: gender-control-categorical
make_wide_categorical_subgroup_table(categorical_control_variables,
                                     category_label_map, "gender_label",
                                     "Categorical Control Variables — By Gender")
```

### By Zone

#### Dependent Variables

```{=latex}
\begin{landscape}
```

```{r}
#| echo: true
#| code-fold: true
#| label: zone-dependent
make_wide_subgroup_table(dependent_variables[1:2], "zone_label",
                         "Dependent Variables — By Zone")
```

```{=latex}
\end{landscape}
```

#### Independent Variables

```{=latex}
\begin{landscape}
```

```{r}
#| echo: true
#| code-fold: true
#| label: zone-independent
make_wide_subgroup_table(independent_variables, "zone_label",
                         "Independent Variables — By Zone")
```

```{=latex}
\end{landscape}
```

#### Control Variables

```{=latex}
\begin{landscape}
```

```{r}
#| echo: true
#| code-fold: true
#| label: zone-control-continuous
make_wide_subgroup_table(continuous_control_variables, "zone_label",
                         "Control Variables — By Zone")
```

```{=latex}
\end{landscape}
```

```{=latex}
\begin{landscape}
```

```{r}
#| echo: true
#| code-fold: true
#| label: zone-control-categorical
make_wide_categorical_subgroup_table(categorical_control_variables,
                                     category_label_map, "zone_label",
                                     "Categorical Control Variables — By Zone")
```

```{=latex}
\end{landscape}
```

### By Residence Type

#### Dependent Variables

```{r}
#| echo: true
#| code-fold: true
#| label: residence-dependent
make_wide_subgroup_table(dependent_variables[1:2], "residence_label",
                         "Dependent Variables — By Residence Type")
```

#### Independent Variables

```{r}
#| echo: true
#| code-fold: true
#| label: residence-independent
make_wide_subgroup_table(independent_variables, "residence_label",
                         "Independent Variables — By Residence Type")
```

#### Control Variables

```{r}
#| echo: true
#| code-fold: true
#| label: residence-control-continuous
make_wide_subgroup_table(
  continuous_control_variables[continuous_control_variables != "residence_type"],
  "residence_label",
  "Control Variables — By Residence Type"
)
```

```{r}
#| echo: true
#| code-fold: true
#| label: residence-control-categorical
make_wide_categorical_subgroup_table(categorical_control_variables,
                                     category_label_map, "residence_label",
                                     "Categorical Control Variables — By Residence Type")
```

## Summary of Key Patterns

::: {.callout-note collapse="true"}
## Key Takeaways from Summary Statistics
:::

```{r}
#| label: plot-setup
#| include: false

library(ggplot2)
library(dplyr)
library(tidyr)

pub_theme <- theme_minimal(base_size = 13) +
  theme(
    plot.title       = element_text(size = 15, face = "bold", hjust = 0.5,
                                    margin = margin(b = 8)),
    plot.subtitle    = element_text(size = 11, hjust = 0.5, color = "grey40",
                                    margin = margin(b = 12)),
    plot.caption     = element_text(size = 9, color = "grey50", hjust = 0,
                                    margin = margin(t = 12)),
    axis.title       = element_text(size = 12, face = "bold"),
    axis.text        = element_text(size = 11),
    legend.title     = element_text(size = 12, face = "bold"),
    legend.text      = element_text(size = 11),
    legend.position  = "bottom",
    legend.key.size  = unit(0.9, "cm"),
    strip.text       = element_text(size = 12, face = "bold"),
    strip.background = element_rect(fill = "#f0f4f8", color = "grey75"),
    panel.grid.major = element_line(color = "grey90"),
    panel.grid.minor = element_blank(),
    panel.spacing    = unit(1.8, "lines"),
    plot.margin      = margin(20, 20, 20, 20)
  )

theme_set(pub_theme)

gender_colors    <- c("Female" = "#e05c5c", "Male" = "#3a7abf")
insurance_colors <- c("No Insurance" = "#e05c5c", "Insured" = "#3a7abf")
pension_colors   <- c("No Pension"   = "#e09a3a", "Pension" = "#3a7abf")
zone_colors      <- c("Central" = "#3a7abf", "East" = "#4aab5e",
                      "North"   = "#e05c5c", "South" = "#9b59b6", "West" = "#e09a3a")
```

# Descriptive Evidence

------------------------------------------------------------------------
```{r caste, include=F}
HH_FI <- HH_FI |>
  mutate(class_caste = case_when(
    caste_social_group ==4 ~ "Upper Caste",
    caste_social_group %in% c(1, 2, 3) ~ "Lower Caste",
    TRUE ~ NA_character_
  ))
```


### Figure 1: Healthcare visit by mpce quintile, to see the economic gradient in healthcare access across economic status groups

```{r}
#| echo: true
#| code-fold: true
#| label: fig1-healthcare-mpce
#| fig-cap: "Figure 1. Any Healthcare Visit by MPCE Quintile."
#| fig-width: 10
#| fig-height: 6

HH_FI |>
  filter(!is.na(mpce_quintile), !is.na(healthcare)) |>
  mutate(
    MPCE = factor(mpce_quintile,
                  levels = c(1, 2, 3, 4, 5),
                  labels = c("Poorest", "Poor", "Middle", "Rich", "Richest"))
  ) |>
  group_by(MPCE) |>
  summarise(Visit = mean(healthcare, na.rm = TRUE) * 100, .groups = "drop") |>
  ggplot(aes(x = MPCE, y = Visit, fill = MPCE)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = paste0(round(Visit, 1), "%")),
            vjust = -0.5, size = 4, fontface = "bold") +
  scale_fill_brewer(palette = "Blues") +
  scale_y_continuous(limits = c(0, 100),
                     labels = function(x) paste0(x, "%"),
                     expand = expansion(mult = c(0, 0.1))) +
  labs(
    title    = "Healthcare Visit by Economic Status",
    x        = "MPCE Quintile", y = "% Visited Any Healthcare Facility",
    caption  = "Source: LASI Wave 1, 2017–18."
  ) +
  theme(legend.position = "none")
```

::: {.callout-note collapse="true"}
## Interpretation — Figure 1
:::


### Figure 2: Healthcare visit by caste class, to see the social stratification in healthcare access across caste groups

```{r}
#| echo: true
#| code-fold: true
#| label: fig2-healthcare-caste
#| fig-cap: "Figure 2. Any Healthcare Visit by Caste Class."
#| fig-width: 8
#| fig-height: 6

HH_FI |>
  filter(!is.na(class_caste), !is.na(healthcare)) |>
  mutate(
    Caste = factor(class_caste, levels = c("Lower Caste", "Upper Caste"))
  ) |>
  group_by(Caste) |>
  summarise(Visit = mean(healthcare, na.rm = TRUE) * 100, .groups = "drop") |>
  ggplot(aes(x = Caste, y = Visit, fill = Caste)) +
  geom_col(width = 0.5) +
  geom_text(aes(label = paste0(round(Visit, 1), "%")),
            vjust = -0.5, size = 5, fontface = "bold") +
  scale_fill_manual(values = c("Lower Caste" = "#E69F00", "Upper Caste" = "#56B4E9")) +
  scale_y_continuous(limits = c(0, 100),
                     labels = function(x) paste0(x, "%"),
                     expand = expansion(mult = c(0, 0.1))) +
  labs(
    title    = "Healthcare Visit by Caste Class",
    x        = "Caste Class", y = "% Visited Any Healthcare Facility",
    caption  = "Source: LASI Wave 1, 2017–18."
  ) +
  theme(legend.position = "none")
```

::: {.callout-note collapse="true"}
## Interpretation — Figure 2
:::


### Figure 3: Healthcare visit by mpce quintile and caste class, to see the intersection of economic status and caste on healthcare access

```{r}
#| echo: true
#| code-fold: true
#| label: fig3-healthcare-mpce-caste
#| fig-cap: "Figure 3. Any Healthcare Visit by MPCE Quintile and Caste Class."
#| fig-width: 12
#| fig-height: 6

HH_FI |>
  filter(!is.na(class_caste), !is.na(mpce_quintile), !is.na(healthcare)) |>
  mutate(
    Caste = factor(class_caste, levels = c("Lower Caste", "Upper Caste")),
    MPCE = factor(mpce_quintile,
                  levels = c(1, 2, 3, 4, 5),
                  labels = c("Poorest", "Poor", "Middle", "Rich", "Richest"))
  ) |>
  group_by(MPCE, Caste) |>
  summarise(Visit = mean(healthcare, na.rm = TRUE) * 100, .groups = "drop") |>
  ggplot(aes(x = MPCE, y = Visit, fill = Caste)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_text(aes(label = paste0(round(Visit, 1), "%")),
            position = position_dodge(width = 0.7),
            vjust = -0.5, size = 3.5, fontface = "bold") +
  scale_fill_manual(values = c("Lower Caste" = "#E69F00", "Upper Caste" = "#56B4E9")) +
  scale_y_continuous(limits = c(0, 100),
                     labels = function(x) paste0(x, "%"),
                     expand = expansion(mult = c(0, 0.1))) +
  labs(
    title    = "Healthcare Visit by Economic Status and Caste Class",
    x        = "MPCE Quintile", y = "% Visited Any Healthcare Facility", fill = "Caste Class",
    caption  = "Source: LASI Wave 1, 2017–18."
  )
```

::: {.callout-note collapse="true"}
## Interpretation — Figure 3
:::

---

### Figure 4: Healthcare Utilization, Insurance, and Pension by Gender

```{r}
#| echo: true
#| code-fold: true
#| label: fig4-descriptive-gender
#| fig-cap: "Figure 1. Healthcare Utilization, Insurance Coverage, and Pension Receipt by Gender. Documenting the gender gap in both utilization and financial protection."
#| fig-width: 13
#| fig-height: 7

HH_FI |>
  filter(!is.na(sex)) |>
  mutate(Gender = factor(sex, 0:1, c("Female", "Male"))) |>
  group_by(Gender) |>
  summarise(
    `Any Healthcare\nVisit`      = mean(healthcare,               na.rm=TRUE)*100,
    `Inpatient\nHospitalization` = mean(inpatient_hospitalization, na.rm=TRUE)*100,
    `Health Insurance\nCoverage` = mean(health_insurance_covered,  na.rm=TRUE)*100,
    `Pension\nReceipt`           = mean(pension_status,            na.rm=TRUE)*100,
    `Financial\nSupport`         = mean(financial_support,         na.rm=TRUE)*100,
    .groups = "drop"
  ) |>
  pivot_longer(cols = -Gender, names_to = "Measure", values_to = "Value") |>
  mutate(
    Category = case_when(
      Measure %in% c("Any Healthcare\nVisit", "Inpatient\nHospitalization") ~
        "Healthcare Utilization",
      TRUE ~ "Financial Protection"
    ),
    Category = factor(Category,
                      levels = c("Healthcare Utilization", "Financial Protection")),
    Measure  = factor(Measure, levels = c(
      "Any Healthcare\nVisit", "Inpatient\nHospitalization",
      "Health Insurance\nCoverage", "Pension\nReceipt", "Financial\nSupport"
    ))
  ) |>
  ggplot(aes(x = Measure, y = Value, fill = Gender)) +
  geom_col(position = position_dodge(width = 0.65), width = 0.55,
           color = "white", linewidth = 0.4) +
  geom_text(aes(label = paste0(round(Value,1), "%")),
            position = position_dodge(width = 0.65),
            vjust = -0.5, size = 3.8, fontface = "bold") +
  facet_wrap(~ Category, scales = "free_x", nrow = 1) +
  scale_fill_manual(values = gender_colors) +
  scale_y_continuous(limits = c(0, 90),
                     labels = function(x) paste0(x, "%"),
                     expand = expansion(mult = c(0, 0.08))) +
  labs(
    title    = "Healthcare Utilization and Financial Protection by Gender",
    x        = NULL, y = "Rate (%)", fill = "Gender",
    caption  = "Source: LASI Wave 1, 2017–18."
  )
```

::: {.callout-note collapse="true"}
## Interpretation — Figure 4

-   About **70% of older women** and **64% of older men** visited any healthcare facility in the past 12 months.
-   But only **6.5% of women** and **7.1% of men** were actually admitted to hospital — meaning most visits were outpatient, not serious enough for hospitalization.
-   Women visit healthcare more than men overall, but men get hospitalized slightly more — suggesting women seek routine care more while men have more serious illness episodes.
-   For financial protection: only **21.6% of women** have health insurance vs **23.8% of men** — women are less covered.
-   Pension is even more unequal — only **3.6% of women** receive pension vs **11.4% of men** — a very large gender gap.
-   **Financial support from family** is received by 15% of women and 11% of men — interestingly women receive more family financial support than men, which is the opposite of the insurance and pension pattern. This suggests that where formal protection (insurance, pension) is lower for women, informal family support partially compensates.
-   This figure sets the **baseline** — it tells us who uses healthcare and who has financial protection before we look at any relationships.
:::

------------------------------------------------------------------------

### Figure 5: How Age Shapes the Insurance Effect on Healthcare Utilization

```{r}
#| echo: true
#| code-fold: true
#| label: fig5-age-insurance
#| fig-cap: "Figure 5. Inpatient Hospitalization by Age Group, Insurance Status, and Gender. It shows whether the insurance effect on hospitalization grows with age and differs by gender."
#| fig-width: 13
#| fig-height: 7

HH_FI |>
  filter(!is.na(age), !is.na(sex), !is.na(health_insurance_covered),
         !is.na(inpatient_hospitalization)) |>
  mutate(
    Age_Group = cut(age, c(44,54,64,74,100),
                    labels = c("45–54","55–64","65–74","75+")),
    Gender    = factor(sex, 0:1, c("Female","Male")),
    Insurance = factor(health_insurance_covered, 0:1, c("No Insurance","Insured"))
  ) |>
  filter(!is.na(Age_Group)) |>
  group_by(Age_Group, Gender, Insurance) |>
  summarise(Hosp = mean(inpatient_hospitalization, na.rm=TRUE)*100,
            .groups="drop") |>
  ggplot(aes(x = Age_Group, y = Hosp, color = Insurance,
             group = Insurance, shape = Insurance)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 5) +
  geom_text(aes(label = paste0(round(Hosp,1), "%")),
            vjust = -0.9, size = 3.5, fontface = "bold", show.legend = FALSE) +
  facet_wrap(~ Gender, nrow = 1) +
  scale_color_manual(values = insurance_colors) +
  scale_shape_manual(values = c("No Insurance" = 16, "Insured" = 17)) +
  scale_y_continuous(limits = c(0, 18),
                     labels = function(x) paste0(x, "%")) +
  labs(
    title    = "How Age Shapes the Insurance Effect on Inpatient Hospitalization",
    x        = "Age Group",
    y        = "% Hospitalized",
    color    = "Insurance Status",
    shape    = "Insurance Status",
    caption  = "Source: LASI Wave 1, 2017–18."
  )
```

::: {.callout-note collapse="true"}
## Interpretation — Figure 5

-Figure 3 now asks: **does this effect grow with age?** - **Yes** — the gap between insured and uninsured hospitalization rates grows larger as age increases.
- At age **45–54** the gap between insured and uninsured is small — younger older adults manage without insurance.
- By age **75+** the gap is much larger — the oldest adults benefit most from insurance coverage.
- This pattern holds for both men and women but is slightly stronger for **women** — older insured women at 75+ have much higher hospitalization rates than uninsured women of the same age.
- This is the most **policy-relevant finding** — it shows that insurance matters most for the very old, who are also the most vulnerable and have the highest healthcare needs.
- Already shown that pension and insurance coverage is already low, this means the people who need protection most — **oldest adults and women** — are the least likely to have it.
:::

------------------------------------------------------------------------

### Figure 6: Pension Receipt and Healthcare Outcomes by Residence and Age

```{r}
#| echo: true
#| code-fold: true
#| label: fig6-pension-outcomes
#| fig-cap: "Figure 6. Association Between Pension Receipt and Inpatient Hospitalization by Age Group and Residence. Line plots show how pension effect on hospitalization varies across age groups in urban vs rural areas."
#| fig-width: 13
#| fig-height: 7

HH_FI |>
  filter(!is.na(age), !is.na(residence_type), !is.na(pension_status),
         !is.na(inpatient_hospitalization)) |>
  mutate(
    Age_Group = cut(age, c(44,54,64,74,100),
                    labels = c("45–54","55–64","65–74","75+")),
    Residence = factor(residence_type, 0:1, c("Urban","Rural")),
    Pension   = factor(pension_status, 0:1, c("No Pension","Pension"))
  ) |>
  filter(!is.na(Age_Group)) |>
  group_by(Age_Group, Residence, Pension) |>
  summarise(Hosp = mean(inpatient_hospitalization, na.rm=TRUE)*100,
            .groups="drop") |>
  ggplot(aes(x = Age_Group, y = Hosp, color = Pension,
             group = Pension, shape = Pension)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 5) +
  geom_text(aes(label = paste0(round(Hosp,1), "%")),
            vjust = -0.9, size = 3.5, fontface = "bold", show.legend = FALSE) +
  facet_wrap(~ Residence, nrow = 1) +
  scale_color_manual(values = pension_colors) +
  scale_shape_manual(values = c("No Pension" = 16, "Pension" = 17)) +
  scale_y_continuous(limits = c(0, 18),
                     labels = function(x) paste0(x, "%")) +
  labs(
    title    = "Pension Receipt and Inpatient Hospitalization by Age Group and Residence",
    subtitle = "Adults aged 45+ | LASI Wave 1, 2017–18",
    x        = "Age Group", y = "% Hospitalized",
    color    = "Pension Status", shape = "Pension Status",
    caption  = "Source: LASI Wave 1, 2017–18."
  )
```

::: {.callout-note collapse="true"}
## Interpretation — Figure 6

-   This figure asks — **does having pension income help older adults access hospital care as they age?**
-   The answer is clearly **yes** — in both urban and rural areas, pension recipients have higher hospitalization rates than those without pension at every age group.
-   In **urban areas** the pension gap is largest at age 45–54 — young urban pensioners (likely formal sector retirees) access hospital care much more than non-pensioned.
-   In **rural areas** the pension effect grows with age — by age 75+ rural pensioners are hospitalized almost **twice as often** as rural non-pensioners.
-   It shows pension strongly increases hospitalization, this means the majority of older adults, especially women, are **missing out on this benefit**.
-   This figure proves- **pension income shaping healthcare behavior** among older adults.
:::

------------------------------------------------------------------------

# Regression Analysis

```{r regression-setup, include=FALSE}
library(haven)
library(dplyr)
library(stargazer)
library(kableExtra)
library(knitr)
library(modelsummary)  
library(fixest)        

options(knitr.table.format = "latex")
options(kableExtra.latex.load_packages = FALSE)

HH_FI <- read_dta("HH_FI.dta")

HH_FI <- HH_FI |>
  mutate(across(where(haven::is.labelled), as.numeric))

HH_FI <- HH_FI |>
  mutate(
    gender_label    = factor(sex, levels = c(0, 1), labels = c("Female", "Male")),
    zone_label      = factor(zone),
    residence_label = factor(residence_type, levels = c(0, 1), labels = c("Urban", "Rural"))
  )

stargazer_output_type <- if (knitr::is_latex_output()) "latex" else "html"
```

## Healthcare Utilization and Insurance Coverage

```{r regressions1, include=T}
#| echo: true
#| code-fold: true

hc_ins_1 <- lm(healthcare ~ health_insurance_covered + factor(state_id), data = HH_FI)

hc_ins_2 <- lm(healthcare ~ health_insurance_covered +
                 age + sex + residence_type +
                 factor(religion) + factor(caste_social_group) +
                 factor(marital_status) +
                 factor(state_id), data = HH_FI)

ip_ins_1 <- lm(inpatient_hospitalization ~ health_insurance_covered + factor(state_id), data = HH_FI)

ip_ins_2 <- lm(inpatient_hospitalization ~ health_insurance_covered +
                 age + sex + residence_type +
                 factor(religion) + factor(caste_social_group) +
                 factor(marital_status) +
                 factor(state_id), data = HH_FI)

```

```{r, results='asis'} 
#| echo: true 
#| code-fold: true


stargazer(hc_ins_1, hc_ins_2, ip_ins_1, ip_ins_2,
          type = "html",
          keep = "health_insurance_covered",
          omit.stat = c("f", "ser", "adj.rsq"),
          add.lines = list(
            c("Controls", "No", "Yes", "No", "Yes"),
            c("State FE", "Yes", "Yes", "Yes", "Yes")
          ),
          table.attr = 'style="width:100%;"')

```


## Healthcare Utilization and Pension Receipt 

```{r regressions2, include=T}
#| echo: true
#| code-fold: true

hc_pen_1 <- lm(healthcare ~ pension_status + factor(state_id), data = HH_FI)

hc_pen_2 <- lm(healthcare ~ pension_status +
                 age + sex + residence_type +
                 factor(religion) + factor(caste_social_group) +
                 factor(marital_status) +
                 factor(state_id), data = HH_FI)

ip_pen_1 <- lm(inpatient_hospitalization ~ pension_status + factor(state_id), data = HH_FI)

ip_pen_2 <- lm(inpatient_hospitalization ~ pension_status +
                 age + sex + residence_type +
                 factor(religion) + factor(caste_social_group) +
                 factor(marital_status) +
                 factor(state_id), data = HH_FI)

```

```{r, results='asis'} 
#| echo: true 
#| code-fold: true
stargazer(hc_pen_1, hc_pen_2, ip_pen_1, ip_pen_2,
          type = "html",
          keep = "pension_status",
          omit.stat = c("f", "ser", "adj.rsq"),
          add.lines = list(
            c("Controls", "No", "Yes", "No", "Yes"),
            c("State FE", "Yes", "Yes", "Yes", "Yes")
          ),
          table.attr = 'style="width:100%;"')
```


## Healthcare Utilization and Financial Support

```{r regressions3, include=T}
#| echo: true
#| code-fold: true
#| tbl-cap: "Effect of Financial Support on Healthcare Utilization"
hc_fin_1 <- lm(healthcare ~ financial_support + factor(state_id), data = HH_FI)

hc_fin_2 <- lm(healthcare ~ financial_support +
                 age + sex + residence_type +
                 factor(religion) + factor(caste_social_group) +
                 factor(marital_status) +
                 factor(state_id), data = HH_FI)

ip_fin_1 <- lm(inpatient_hospitalization ~ financial_support + factor(state_id), data = HH_FI)

ip_fin_2 <- lm(inpatient_hospitalization ~ financial_support +
                 age + sex + residence_type +
                 factor(religion) + factor(caste_social_group) +
                 factor(marital_status) +
                 factor(state_id), data = HH_FI)
```
```{r, results='asis'} 
#| echo: true
#| code-fold: true
stargazer(hc_fin_1, hc_fin_2, ip_fin_1, ip_fin_2,
          type = "html",
          keep = "financial_support",
          omit.stat = c("f", "ser", "adj.rsq"),
          add.lines = list(
            c("Controls", "No", "Yes", "No", "Yes"),
            c("State FE", "Yes", "Yes", "Yes", "Yes")
          ),
          table.attr = 'style="width:100%;"')
```

